---
title: "03_augment"
format: html
---

### Content ideas - enhanced dataset

-   If needed and relevant, categorize blood pressure into low, normal, high, very high

-   summary staticstics to begin the analysis

-   Create age groups

# Data augmentation

Here, we modify the cleaned dataset to prepare it for analysis. As a first step, we simply categorize the ages into age groups by intervals of 10, and re-order the data such that the new `age_group` column is after `age` for improved structure.

```{r create_age_groups}
kidney_augmented <- kidney_clean |> 
  mutate(age_group = case_when(age <= 10 ~ "0-10",
                               age > 10 & age <= 20 ~ "11-20",
                               age > 20 & age <= 30 ~ "21-30",
                               age > 30 & age <= 40 ~ "31-40",
                               age > 40 & age <= 50 ~ "41-50",
                               age > 50 & age <= 60 ~ "51-60",
                               age > 60 & age <= 70 ~ "61-70",
                               age > 70 & age <= 80 ~ "71-80",
                               age > 80 & age <= 90 ~ "81-90",
                               age > 90 & age <= 100 ~ "91-100")) |> 
  relocate(age_group,
           .after = age)
```

Next, we split the dataset into two dataframes:

-   One with only the diagnoses (`hypertension`, `coronary_artery_disease`, `anemia`, and `classification)`

-   One with the remaining vitals and age information.

```{r split_data}
diagnoses = c("hypertension",
              "coronary_artery_disease",
              "anemia",
              "classification")

```
