---
title: "05_analysis_1"
format:
  html:
    embed-resources: true
    output-file: 05_analysis_1.html
    output-dir: results
editor: visual
execute: 
  echo: true
  message: false
  warning: false
---

## Loading the data

```{r}
library(tidyverse)

kidney_merged_GFR <- read_rds("../data/03_data_augment.rds")

```

## Prepare variables for Analysis

Here we make sure that the key categorical variables are treated as factors and that the CKD stages have an ordered structure.

```{r}
kidney_analysis <- kidney_merged_GFR |>
mutate(
  classification = as.factor(classification),
  hypertension = as.factor(hypertension),
  diabetes_mellitus = as.factor(diabetes_mellitus),
  coronary_artery_disease = as.factor(coronary_artery_disease),
  anemia = as.factor(anemia),
  pedal_edema = as.factor(pedal_edema),
  cdk_stage = factor(
    cdk_stage,
    levels = c("stage_1", "stage_2", "stage_3a", "stage_3b", "stage_4", "stage_5"),
    ordered = TRUE
    )
  )
```

## Question 1: Is kidney function (GFR) lower in CKD patients?

We expect individuals with a CKD diagnosis to have lower estimated glomerular filtration rate (GFR).

```{r}
kidney_analysis |>
  drop_na(GFR_average, classification) |>
  group_by(classification) |>
  summarise(
    n = n(),
    mean_GFR = mean(GFR_average),
    median_GFR = median(GFR_average),
    sd_GFR = sd(GFR_average)
    )

```

Visualize GFR by CKD classification

```{r}
kidney_analysis |>
  drop_na(GFR_average, classification) |>
  ggplot(aes(x = classification, y = GFR_average, fill = classification)) +
  geom_boxplot() +
  theme_minimal() +
<<<<<<< HEAD
=======
  theme(legend.position="none") +
>>>>>>> 4a94c2d6425f07ecd82892108f68c0d82f1e4458
  labs(
    title = "Estimated GFR by CKD classification",
    x = "Classification",
    y = "Estimated GFR (ml/min)"
    ) +
<<<<<<< HEAD
  scale_fill_manual(
    values = c("ckd" = "#EE3B3B", "notckd" = "#1b9e77"),
    labels = c("ckd" = "Diagnosed", "notckd" = "Not diagnosed")
    )
=======
  scale_x_discrete(labels = c("CKD diagnosis", "No CKD diagnosis"))
>>>>>>> 4a94c2d6425f07ecd82892108f68c0d82f1e4458

```

## Question 2: How are CKD stages distributed between diagnosed and non-diagnosed individuals?

Here we look at how the GFR-based CKD stages are distributed in people with and without a CKD diagnosis.

```{r}
kidney_analysis |>
  drop_na(cdk_stage, classification) |>
  count(cdk_stage, classification) |>
  group_by(classification) |>
  mutate(prop = n / sum(n))

```

```{r}
stage_labs <- c("Normal", 
                "Mild", 
                "Mild to moderate", 
                "Moderate to severe",
                "Severe", 
                "Failure")

kidney_analysis |>
  drop_na(cdk_stage, classification) |>
  ggplot(aes(x = cdk_stage, fill = classification)) +
  geom_bar(position = "fill") +
  theme_minimal() +
<<<<<<< HEAD
  labs(
    title = "Distribution of GFR-based CKD stages by classification",
    x = "CKD stage (based on GFR)",
    y = "Proportion within classification group"
    ) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(
    values = c("ckd" = "#EE3B3B", "notckd" = "#1b9e77"),
    labels = c("ckd" = "Diagnosed", "notckd" = "Not diagnosed")
    )
=======
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Distribution of GFR-based CKD stages by classification",
    x = NULL,
    y = "Proportion within classification group"
    ) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_discrete(
    labels = c("ckd" = "CKD diagnosis", "notckd" = "No CKD diagnosis")
    ) +
  scale_x_discrete(labels = stage_labs)
>>>>>>> 4a94c2d6425f07ecd82892108f68c0d82f1e4458

```

## Question 3: Are hypertension and diabetes more common among CKD patients?

Hypertension and diabetes are known risk factors for CKD. Here we compare their prevalence between groups.

```{r}
kidney_analysis |>
  drop_na(hypertension, classification) |>
  count(classification, hypertension) |>
  group_by(classification) |>
  mutate(prop = n / sum(n))

```

Visualization

```{r}
kidney_analysis |>
  drop_na(hypertension, classification) |>
  ggplot(aes(x = hypertension, fill = classification)) +
<<<<<<< HEAD
  geom_bar(position = "dodge") +   
=======
  geom_bar(
    position = position_dodge(preserve = "single")) +   
>>>>>>> 4a94c2d6425f07ecd82892108f68c0d82f1e4458
  theme_minimal() +
  labs(
    title = "Hypertension by CKD classification",
    x = "Hypertension",
    y = "Number of people"
  ) +
<<<<<<< HEAD
  scale_fill_manual(
    values = c("ckd" = "#EE3B3B", "notckd" = "#1b9e77"),
    labels = c("ckd" = "Diagnosed", "notckd" = "Not diagnosed")
  )
=======
  scale_fill_discrete(
    labels = c("ckd" = "CKD diagnosis", "notckd" = "No CKD diagnosis")
    )
>>>>>>> 4a94c2d6425f07ecd82892108f68c0d82f1e4458

```

## Diabetes vs Classification

```{r}
kidney_analysis |>
  drop_na(diabetes_mellitus, classification) |>
  count(classification, diabetes_mellitus) |>
  group_by(classification) |>
  mutate(prop = n / sum(n))

```

```{r}
kidney_analysis |>
  drop_na(diabetes_mellitus, classification) |>
  ggplot(aes(x = diabetes_mellitus, fill = classification)) +
<<<<<<< HEAD
  geom_bar(position = "dodge") +   # <-- side by side count comparison
=======
  geom_bar(
    position = position_dodge(preserve = "single")) +
>>>>>>> 4a94c2d6425f07ecd82892108f68c0d82f1e4458
  theme_minimal() +
  labs(
    title = "Diabetes mellitus by CKD classification",
    x = "Diabetes mellitus",
    y = "Number of people"
  ) +
<<<<<<< HEAD
  scale_fill_manual(
    values = c("ckd" = "#EE3B3B", "notckd" = "#1b9e77"),
    labels = c("ckd" = "Diagnosed", "notckd" = "Not diagnosed")
  )


=======
  scale_fill_discrete(
    labels = c("ckd" = "CKD diagnosis", "notckd" = "No CKD diagnosis")
    )
>>>>>>> 4a94c2d6425f07ecd82892108f68c0d82f1e4458
```
